Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-11-15 paper/paper.tex
adjusted supplements to figure 2

\caption{
Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\FIGSUPP[workflow]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
{\bf (C)}
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

\figsupp[MACS enrichment of IFN+ cells.]
{
Example MACS enrichments of IFN+ influenza-infected cells.
A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
{\bf (A)}
After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment in IFN+ cells.
We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
{\bf (B)}
Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
The qPCR was performed in quadruplicate (hence the four points for each sample).
}
{\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}}
\label{figsupp:MACS}

\figdata{Genbank files giving sequences of the wild-type and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
\label{figdata:virus_seqs}

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\FIGSUPP[workflow]{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

@@

\subsection{Viruses for single-cell experiments}
We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
We used both the wild-type virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \FIGDATA[workflow]{virus_seqs}.
These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.

@@

The wild-type and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
The cells were then infected with this mixture at a dose designed to infect about half the cells (\FIG{transcriptomics}C suggests that the actual rate of detectable infection was slightly lower).
Infections were allowed to proceed for 12 hours.
The cells were then trypsinized, the trypsin was quenched with D10 media, and cells were resuspended in de-gassed PBS supplemented with 0.5\% bovine serum albumin and 5 mM EDTA. 
To enrich IFN+ cells, the cells were then incubated with anti-LNGFR MACSelect Microbeads (Miltenyi Biotec) and twice passed over an MS magnetic column (Miltenyi Biotec), retaining the bound (and presumably IFN-enriched) population each time. 
This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in \FIGSUPP[workflow]{MACS}.
The original, unsorted, population was then added back in to $\sim$10\% of the final cell fraction in order to ensure the presence of interferon negative cells. 
At this point, uninfected canine (MDCK-SIAT1) cells were also added to $\sim$5\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{transcriptomics}A) and background viral mRNA in uninfected cells (\FIG{transcriptomics}C). 
We began this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.

@@

Briefly, the raw deep sequencing data from the Illumina 3'-end sequencing were processed using the 10X Genomics software package \texttt{cellranger} (version 2.2.0). 
We built a multi-species alignment reference consisting of a concatenation of the human and influenza virus transcriptomes (the first ``species'') and the canine transcriptome (the second ``species''). 
The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \FIGDATA[workflow]{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (\FIG{transcriptomics}A), and generate a matrix giving the expression of each gene in each single cell.
We used a custom Python script to determine the number of influenza virus reads that could be assigned to the wild-type or synonymously barcoded virus, and added this information to the annotated the cell-gene matrix.

@@

@@

>>>>

\caption{
Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
{\bf (C)}
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

@@

\subsection{Viruses for single-cell experiments}
We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
We used both the wild-type virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.

@@

The wild-type and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
The cells were then infected with this mixture at a dose designed to infect about half the cells (\FIG{transcriptomics}C suggests that the actual rate of detectable infection was slightly lower).
Infections were allowed to proceed for 12 hours.
The cells were then trypsinized, the trypsin was quenched with D10 media, and cells were resuspended in de-gassed PBS supplemented with 0.5\% bovine serum albumin and 5 mM EDTA. 
To enrich IFN+ cells, the cells were then incubated with anti-LNGFR MACSelect Microbeads (Miltenyi Biotec) and twice passed over an MS magnetic column (Miltenyi Biotec), retaining the bound (and presumably IFN-enriched) population each time. 
This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in \SUPPFIG{MACS}.
The original, unsorted, population was then added back in to $\sim$10\% of the final cell fraction in order to ensure the presence of interferon negative cells. 
At this point, uninfected canine (MDCK-SIAT1) cells were also added to $\sim$5\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{transcriptomics}A) and background viral mRNA in uninfected cells (\FIG{transcriptomics}C). 
We began this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.

@@

Briefly, the raw deep sequencing data from the Illumina 3'-end sequencing were processed using the 10X Genomics software package \texttt{cellranger} (version 2.2.0). 
We built a multi-species alignment reference consisting of a concatenation of the human and influenza virus transcriptomes (the first ``species'') and the canine transcriptome (the second ``species''). 
The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \SUPPFILE{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (\FIG{transcriptomics}A), and generate a matrix giving the expression of each gene in each single cell.
We used a custom Python script to determine the number of influenza virus reads that could be assigned to the wild-type or synonymously barcoded virus, and added this information to the annotated the cell-gene matrix.

@@

\begin{suppfig}
\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}
\caption{
Example MACS enrichments of IFN+ influenza-infected cells.
A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
{\bf (A)}
After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment in IFN+ cells.
We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
{\bf (B)}
Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
The qPCR was performed in quadruplicate (hence the four points for each sample).
}
\label{suppfig:MACS}
\end{suppfig}

@@

\begin{suppfile}
\caption{
Genbank files giving sequences of the wild-type and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
\label{suppfile:virus_seqs}
\end{suppfile}
